\documentclass[authoryear,review,3p]{elsarticle}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage[normalem]{ulem}

\usepackage{lineno,hyperref}
\usepackage{amsmath,amsfonts,amssymb}

\usepackage{listings}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{multirow}
\usepackage{subfigure}
\usepackage{hyperref}
\usepackage[acronym,nomain]{glossaries}

\begin{document}

\begin{frontmatter}

\title{Statistical Learning of Biological Structure in the Human Brain}

\author{Dr. med. Danilo Bzdok}

\end{frontmatter}

\bigskip
\bigskip
\bigskip
\centerline{
"But above all, master technique and produce original data; 
all the rest will follow."}
\centerline{Santiago Ram\'{o}n y Cajal}

\bigskip
\bigskip
\bigskip

\textbf{Supervisors\\}

Prof. Dr. rer.-nat. Stefan Conrad, Natural Science Faculty, HHU D\"usseldorf, Germany\\

Prof. Dr. med. Simon Eickhoff, Medical Faculty, HHU D\"usseldorf, Germany\\

Dr. Bertrand Thirion, INRIA, Saclay, France\\


Send for review:
Elvis, Michael, Guillaume, Denis, Conrad, Robert,
Bertrand, JR, Konstantin


\bigskip

\newpage
\section*{Publications related to the present dissertation}

\linebreak
Cumulative Impact Factor: $\approx$60

\linebreak
\textit{Original papers}

\textbf{Bzdok D}, Grisel O, Eickenberg M, Thirion B, Varoquaux G.
Semi-supervised Factored Logistic Regression for High-Dimensional
Neuroimaging Data. Under review at NIPS.

\textbf{Bzdok D}, Grisel O, Eickenberg M, Varoquaux G, Poupon C, Thirion B.
Network-network architecture: Generative models of task activity patterns.
Under review at Cerebral Cortex.

Bludau S*, \textbf{Bzdok D*}, Gruber O,
Kohn N, Riedl V, Müller V, Hoffstaedter F, Eickhoff SB.
Medial prefrontal aberrations in major depressive disorder
revealed by cytoarchitonically informed voxel-based morphometry.
\textit{American Journal of Psychiatry}, in press. *equal contributions

\textbf{Bzdok D*}, Hartwigsen G*, Reid A, Eickhoff SB.
A hierarchy of the left inferior parietal lobe in social cognition and
language.
\textit{Neuroscience and Biobehavioral Reviews}, in press. *equal contributions

\textbf{Bzdok D}, Heeger A, Langner R, Laird A, Fox P, Palomero-Gallagher,
Vogt BA, Zilles K, Eickhoff SB.
Subspecialization in the human posterior medial cortex.
\textit{Neuroimage}, in press.

Eickhoff SB, Laird AR, Fox PT, \textbf{Bzdok D}*, Hensel L*.
Functional segregation of the human dorsomedial prefrontal cortex.
\textit{Cerebral Cortex}, in press. *equal contributions

\textbf{Bzdok D}, Langner R, Schilbach L, Laird AR, Fox PT, Zilles K, Eickhoff SB.
Characterization of the temporo-parietal junction by combining data-driven
parcellation, complementary connectivity analyses, and functional decoding.
\textit{Neuroimage}, 2013.

\bigskip
\textit{Review and opinion papers}

Eickhoff SB, Thirion B, Varoquaux G, \textbf{Bzdok D}.
Connectivity-based parcellation: critique \& implications.
\textit{Human Brain Mapping}, in press.

Eickhoff, SB \& \textbf{Bzdok D}.
Neuroimaging and modeling. Where is the road to clinical application?
\textit{Der Psychiater}, 2014, in press. 

Eickhoff SB \& \textbf{Bzdok D}.
[Statistical meta-analyses in imaging neuroscience.]
\textit{Klinische Neurophysiologie}, 2013, 44:199-203.

\bigskip
\textit{Book chapters}

\linebreak
\textbf{Bzdok D} \& Eickhoff SB.
Statistical learning of the neurobiological of schizophrenia.
In: \textit{The neurobiology of schizophrenia}, Springer, Heidelberg.

\newpage

\section*{1 Introduction}

\subsection*{1.1 Analytical and heuristic accesses to nature}

The world around us is complex and volatile.
%
A large proportion of human research efforts are undertaking
in an \textit{analytical} fashion
based on the 
"the unreasonable effectiveness of mathematics in the natural sciences."
This was phrased by the Hungarian-American
physicist, mathematician, and Nobel laureate Eugene P. Wigner (1960).
The language of mathematics is a powerful tool to
describe, formalize, and predict phenomena in nature.
The author emphasizes that it is not imperative that
natural regularities exist in the world. He goes on to
say that it might be even more surprising that humans can actually
find these regularities and use them to their advantage.
Similarly, Albert Einstein said:
"The most incomprehensible, is that the world is comprehensible." 
%
Starting from human-conceived axiomes
we have derived always more complicated
properties of and relationships between mathematical objects by formal proofs
(Connes A., "A view of mathematics").
A logical pyramid of theoreoms is built that lead to always
more general asserations.
We also have detailed knowledge of the limitations of these mathematical
assertions.
%
On the one hand,
an identical regularity can often be equally well described in very distant
branches of mathematics.
On the other hand,
identical mathematical conclusions have reemerged from derivation of
a priori unrelated assertions.
%
Indeed, the same formal language has proofed very apt in
the study of completely unrelated topics and diverging scientific disciplines.
This includes movements of celestial objects in the universe
studied in astronomy
and
the metabolism pathways governing the inner life of the cell
studied in biochemistry.
%
Many rules about the world can thus be perfectly grasped
(Hardy, "Apology").
As another simple example,
Fibonacci numbers (1, 1, 2, 3, 5, 8, 13, etc.)
reappear in many natural phenomena.
The number of petals of a flower and the spirals of a pinapple tend
to be Fibonacci sequences.
The family tree of honey bees is also governed by Fibonacci regularities.
Even the proportions of human finger bones follow this formalism.
Knowledge of such mathematical regularities
allows to impose logical structure on the external world.
%
It remains an unresolved philosophical debate whether we have
\textit{discovered} or \textit{invented} mathematics.
Yet, there is probably no doubt
that mathematical conceptualization
evolves as a feature of human cultural evolution (Tomasello, 2001).
Even the most abstract mathematical concepts can
be exchanged between individuals. Consequently,
this knowledge resource can be easily passed
on across generations and geographical distances.
One may note that
there is usually consensus among mathematicians
about the architecture of their discipline.
From an anthropological perspective,
mathematical formalism
appears to be one of the most
powerful tools and most defining properties of the human species
(S. Dehaene, "The Number Sense").
Indeed,
Eugene Wigner concludes his praise of equations with the following words:
"The miracle of the appropriateness of the language of mathematics
for the formulation of the laws of physics is a wonderful gift
which we neither understand nor deserve. We should be grateful for
it and hope that it will remain valid in future research [...]."
(1960, p. 14).


However, this dogma has repeatedly been challenged formally
and empirically.
%
In formal approaches,
mathematics were shown incomplete and inherently contradictory because,
in any axiomatic systems, some true assertions cannot be proven
("incompleteness theorems", G\"odel, 1931).
Additionally,
it is possible to define a real number with equidistributed
digits that can however not be computed
("Chaitin's constant $\Omega$", Chaitin, 2006).
%
In empirical approaches,
the omnipotence of mathematical equations has been challenged
as the
best possible way to describe and predict nature
(Halevy et al., 2009; Hinton/LeCun/Bengio, 2015; Pietsch, 201?).
%
This shift in the scientific discourse is made explicit by
three recent empirical observations:
\begin{enumerate}
  \item Sophisticated, more accurate models can be outperformed by
  simple models that are fit with massive training data.
  \item Simple models trained with rich data can outperform
  models that have been designed according to
  extensive domain expertise.
  \item There appears to be a minimal threshold of training data
  for deriving models to suddenly exhibit emergence properties.
\end{enumerate}

One prominent example for modelling nature in such a \textit{heuristic} fashion is
automatic language translation of human text and speech.
Translation systems based on 
human-derived systems of grammar rules have perhaps never achieved
satisfactory success (P. Norvig, 2011, "On Chomsky").
That is, analytical approaches based on
thousands of books pages storing domain expertise in form of
deterministic rules appeared insufficient for building
language models that can cope with real-world settings.
%
Statistical machine translation was more successful
by implementing probabilistic hidden Markov models (HMM)
as a heuristic approach.
%
The next word or N-gram is predicted
only by the one (order 1) or few (order n) preceding ones
with equal transition probabilities (Bengio, 199X "Markovian Models").
This special case of a recurrent neural network
computes the conditional probabilities of the next language element
depending on the most recent history of these elements
based on a dictionary of known elements.
The transition matrix of human language can be easily \textit{learned}
from data by observing a preferably long stream of
real-world language (i.e., a "corpus").
%
The class of HMMs
thus became a dominanting feature
of computational linguistics.
%
Today, virtually all professional translation software solutions
for both written and spoken language
are enabled by heuristic mathematical models.

As an example from human biology,
we currently have only few means to predict the
toxicity of environmental chemicals and
potential effects of new drug compounds on health.
%
The complex and unknown phenonon in nature here pertains
to the causal link from
a protein's \textit{known} primary structure
(i.e., 1D string of aminoacids)
to
its \textit{unknown} tertiary/quaternary structure
(i.e., the combination of 3D foldings
of the aminoacid string that subserves function).
%
Among the millions of existing proteins
we only know the tertiary/quaternary structure of
approximately 30.000 ones.
The structural configuration is however necessary to identify the position of
bindings sites for protein-protein interaction.
Knowledge of these sites in 3D space is crucially important
to infer that protein's interplay with the human body.
%
In this case,
the learning problem is to
derive a computational model from
massive pairs of known primary protein structure and
known protein properties, including toxicity,
by intentionally treating the 3D structure
as a black box.
%
State-of-the-art neural network algorithms
have very recently solved this heuristic learning problem
better than any previous approach in academia or industry
(Dahl et al., 2014; Unterthiner et al., 2015).
These investigators thus showed that biological activity
can be reliably predicted from single aminoacid strings
even without recourse to biological domain expertise.
%
There might currently not exist
an analytical counterpart to such
structure-function mapping of proteins.
%
%
Similar probabilistic scenarios would include predicting
political outcomes,
optimizing advertisement strategies,
algorithmic trading in stock markets, and
controlling self-driving cars.


In sum,
humans can create machines to derive
algorithmic predictions from the data.
Observing a phenomenon in nature a sufficient number of times
might be sufficient to
algorithmically extract
the heuristics of its interaction behavior with the world.
This has recently entailed a shift
of attention
from model complexity to data complexity
and
from purely mathematical treatment to giving up some control to self-emergening patterns.
In the absence of an analytical alternative access,
simple mathematical models can thus automatically
formalize diverse classes of natural phenomena
depending on the quality and quantity of
the available data resources.



\subsection*{1.2 Two cultures of statistical modelling}

Statistics is a branch of mathematics that has arguably been the
overall most successful information science.
Statistics aim at extracting information from data
about the mechanisms in nature that generated these data.
%
Given its eclectic character, it may come as no surprise that statistics
has developed both analytical and heuristic strategies
to model regularities of phenomena in nature.
Yet, analytical and heuristic statistical cultures
have developed independently (Breiman, 2011).
They differ with regard to historical origin, mathematical foundation,
and modelling goal.

The overwhelming majority of statisticians
follow an analytical regime by
adhering to \textit{classical statistics} (CS) for
\textit{data modelling}.
They hold that the phenomenon under study can be viewed as a black box
whose inner workings can be described by a small set of
underlying variables.
It is up to the statistician in charge
to choose the model that best reflects nature.
Data are then used to estimate the parameters of that pre-specified model.
Classical statistics have dominated research at the universities
for almost 90 years now.
%
Well known members of the CS family include for instance
Student's t-test, ANOVA,
and Chi-squared test.
\textit{Statistical hypothesis testing} has been introduced in the beginning
of the last century (Fisher, 1925; Neyman and Pearson, 1928).
The same approach is still practiced today in its original form (Goodman, 1999).  
%
The ensuing \textit{p-value} measures how likely it is
to observe the data at hand
assuming the non-preferred null hypothesis ($H_0$)
to find indirect evidence
for the preferred alternative hypothesis ($H_1$).
%
Despite the prevailing presence of p-values,
it has not been conceived by Fisher as an acid test
to judge existing versus non-existing effects in nature.
Rather, the intention was a preliminary tool to
filter which potential effects should be more explicitly tested (Nuzzo, 2014).
%
Notably, the drawn conclusions may be wrong
if the hand-selected model is a bad description of
the natural phenomenon under study.
%
Nevertheless, statistical hypothesis testing probably fit perfectly
in its time of inception and adoption.
Indeed, it was designed for use with mechanical calculators
(Efron and Tibshirani, 1991). Gaussian distributional assumptions
have been very useful in many instances for
mathematical convenience and, hence, computational tractability.
Additionally, it suited perfectly the Popperian view of
critical empiricism in academic discourse
(Popper, 1935/2005 "Logik der Forschung"):
scientific progress is to be made by continuous replacement of current
hypotheses by always more explanatory hypotheses
by means of \textit{verification} and \textit{falsification}.
The rationale behind hypothesis falsification
is that even a lot of evidence cannot confirm
a given theory in an \textit{inductive} way, 
while a single counter example is able to proof a theory wrong
in a \textit{deductive} way.
%
In sum,
classical statistics was mostly fashioned
for problems with few data points that can be modeled 
by plausible models with a small number of parameters chosen by the
investigator.


In contrast, only a small minority of statisticians
follow a heuristic regime by
adhering to \textit{statistical learning} (SL) for
\textit{algorithmic modelling}.
This statistical framework is frequently adopted by computer scientists,
physicists, ingenieurs, and others without formal statistical brackground
that are typically working in industry
rather than academia (cf. Daniel and Wood, 1971).
%
They hold that natural phenomena
can be studied by estimating regularities in the inputs and
outputs to the black box without making assumptions
about its internal "true" mechanisms.
A statistical model is thus derived that expresses
relationships between the input and output variables
whose parameters are learned by training data (Abu-Mostafa, 2012).
Put differently, a new function with potentially thousands of
parameters is created
that can predict the output from the input alone,
without explicit programming model.
The input data thus need to represent different variants of
all relevant configurations of the examined phenomenon in nature.
Well-known members of
the SL family include for instance k-means clustering,
Lasso/Ridge regression, and support vector machine classification.
%
Please note that SL here summarizes the seemingly more specific terms
"data-mining", "pattern recognition", "artificial intelligence",
and "machine learning" that are often employed inconsistently.
The independent historical origin of CS and SL families is even
witnessed by the most basic terminology. 
In the CS literature inputs to statistical modeling
are classically called \textit{independent variables}
or, more recently, \textit{predictors},
while these are commonly referred to as \textit{features}
in the SL literature (Hastie et al., 2011).
%
When evaluating whether a certain problem is a possible target for SL
three requirements come into play (Abu-Mostafa, 2012):
\begin{enumerate}
  \item A regularity exists
(if there is no pattern, then it might still be worth trying SL).
  \item The regularity cannot be formalized analytically
(otherwise one can still apply SL, but it might not create the best model).
  \item We have data on the problem (the more, the better).
\end{enumerate}
%
This statistical framework led to a surge of new
computer-intensive statistical techniques since 1980
that can be difficult to compute on a normal calculator
and
that are less concerned with mathematical tractability (Efron, 1991).
In the 95'ies suppor vector machines (SVMs)
proofed to be better than the back then de-facto-standard neural networks
in both classification and regression problems (Valpnik, 1996).
This development has been flanked by changing properties of datasets that
are always higher-dimensional (i.e., more features per observation)
and
based
on larger samples (i.e., more observations).
This is a trend that is not specific to
neuroimaging research but also takes places
in other scientific disciplines,
including but not exclusive to weather forecasting and economic predictions
(Manyika et al., 2011).
In sum,
statistical learning was mostly fashioned
for problems with many data points with largely unknown
data generating processes
that are emulated by a mathematical function
created en passant by a machine.


Importantly, some statistical
methods cannot be easily categorized by the CS-SL distinction.
Statistical methods do, in fact, span a continuum between the two poles of CS and SL
(Jordan/Frontiers in Massive Data, p. 61).
Nevertheless, the two families of statistical methods
can be easily distinguished by a number of archetypical properties.
Bayesian statistics are however orthogonal to the CS-SL distinction
and can be adopted in both methodological families in various flavors.
%
Neither can the terms univariate versus multivariate
(i.e., relying on one versus more than one input variable)
be clearly grouped into either CS or SL.
%
More generally,
neither CS nor SL can generally be considered superior.
This is captured by the \textit{no free lunch theorem}
stating that no single statistical strategy can#
consistently do better in all circumstances (Domingos, 2012; Wolpert, 1996).
The challenge relies in choosing
the statistical approach that is best suited
to the neurobiological phenomenon under study and the neuroscientific research object at hand.


Regarding modelling goals, CS and SL exhibit various differences.
CS typically aims at modeling the black box by making a set of
accurate assumptions about its content,
e.g. the type of signal distribution.
Contrarily, SL typically aims at finding any way to model
the output of the black box from its input
while making the least assumptions possible (Abu-Mostafa et al., 2012).
In CS the phenomenon is therefore treated as partly known
(i.e., the stochastic processes that generated the data),
whereas in SL the phenomenon is treated as complex,
completely unknown, and partly unknowable.
It is in this way that CS tends to be
analytical
(i.e., imposing mathematical rigor on the phenomenon),
whereas SL tends to be
heuristic
(i.e., finding useful approximations to the phenomenon).
CS assumes a given statistical model at the beginning of the investigation,
whereas in SL the model is
generated in the process of the statistical investigation.
In more formal terms,
CS therefore closely relates to parametric statistics
for \textit{confirmatory} data analysis,
whereas SL closely relates to non-parametric statistics
for \textit{exploratory} data analysis
(Tukey, 1977 "Exploratory data analysis").
In more practical terms, CS is typically applied to experimental data
that were generated
the investigator controlled the variables of interest 
(i.e., the system under studied is perturbed),
while SL is typically applied to observational
data without such structured influence by the investigator
(i.e., the system is left unperturbed) (Domingos, 2012).
The work unit for CS is the quantified
significance associated with a statistical
relationship between few variables given a pre-specified model.
The work unit for SL is the quantified robustness of patterns
between many variables or, more generally,
the robustness of \textit{special structure} in the data (Hastie et al., 2011).
CS therefore tests for a particular structure in the data,
whereas SL explores and discovers structure in the data.
Formally, CS implements data modeling by
imposing an a priori model in a top-down manner,
whereas SL implements algorithmic modeling by fitting
a model as a function of the data at hand in a bottom-up manner.
%
Intuitively, the "truth" is believed
to be in the model (cf. Wigner, 1960) in a CS-contrained world,
while it is believed to be in the data
(cf. Halevy et al., 2009) in a SL-constrained world.


As a drastically oversimplified, yet useful, conclusion,
CS preassumes and tests \textit{a model for the data},
whereas SL learns \textit{a model from the data}.
%
Indeed,
both human and computer learning are theoretically
more conceivable in a probabilistic rather than
deterministic sense
(Abu-Mostafa et al., 2012; Dayan et al., 1995; Friston, 2010; Gregory, 1980).
Moreover, each probabilistc model can be viewed as a superclass
of a deterministic model (P. Norvig, "On Chomsky").
%
Taken together,
CS assumes that the data behave according to known mechanisms,
whereas SL exploits
computer algorithms to avoid the a-priori
specifications of data mechanism.



\subsection*{1.3 The human brain as a complex phenomenon in nature}
The human brain is a prime example of
a black box that is complex, mysterious, and perhaps in part unknowable.
It is frequently proposed that
the human brain might be the most complex object in the known universe
(Nature editorial, october 2014).
With the language from above,
the human brain might constitute a phenomon in nature that
can \textit{not} be perfectly grasped by mathematical formalism alone.
More concretely,
the \textit{most pertinent structure}
that we should assume for the human brain,
when measured by
contemporary functional neuroimaging techniques (cf. next passages),
is currently unknown.
Hence, the neuroimaging access to neuroscience can readily be framed as
a problem of \textit{representation learning} (Bengio, 2014).
It is conceivable
that this task can be solved without exhaustive
multi-level neurobiological knowledge (Bostrom, 2014).
This is always more supported by empirical evidence
(e.g., Helmstaedter, M. et al. "Connectomic reconstruction" 2013 Nature)
and
it is a contention that is embraced by the present disseration.


From a global perspective,
the genetic difference between our genetic equipment and that
of our closest ancestors, the non-human primate, turns out to be 
strikingly small.
This has encouraged the conviction that one or very few key genetic
adaptations in the primate lineage have unchained an avalanche
of cognitive and cultural inventions that led up to today's civilization
(Tomasello, 2001).
That is, the human species might be much more defined by the
increasingly fast cultural evolution rather the ramifications
of slow biological evolution.
Crucial cognitive improvements,
such as the emergence of verbal language, 
might have fueled cultural improvements that, in turn, enabled
further cognitive improvements et cetera pp.
This form of \textit{online learning} is a very plausible and decisive
property of intact tissue of the central nervous system.
%
As a first challenge in brain science,
it might therefore be impossible to cleanly dissect
the nature-nurture interplay into independent contributing factors that act
during
phylogeny (i.e., development of the species)
and
ontogeny (i.e., development of an individual organism).
%
In this sense,
investigating the limits between "nature" and "culture" in humans
might equate with asking a paradoxical question
(Dehaene \& Cohen, 2007).
%
Instead,
a necessary factor for the high level of abstraction in human culture
might have precisely been the inextricability, due to bidirectional influence, of
neurobiological plasticity and relentless cultural exchange
between human individuals
in a non-stop, autopoietic optimization process
(Vygotsky 1978, "Mind in Society"; Luhmann 1984, "Soziale Systeme";
Bengio 2013, "Evolving Culture").


Given this recent acceleration in cultural evolution
(cf. Paul Virilio, "Open Skype"),
it might be
rather unlikely that the human brain has developed dedicated
neuronal populations to subserve the panoply of novel behaviors.
Rather, evolutionarily recent mental skills
(e.g., reading and writing, explicit pedagogy, and
symbolic mathematics)
are realized by recombining low-level circuits that initially
developed for other functional roles.
This view has become known as
"neural reuse" and "neural recycling" hypotheses
(Anderson, 2010; Dehaene \& Cohen, 2007).
Non-human primates are lacking many of the sophisticated
mental operations that
are crucially important for maintaining human societies
(Mesulam, 1998; Tomasello, 2003).
In fact,
the "social brain hypothesis" states that our
computationally powerful brains are not an adaptation to
solve problems posed by the physical environment,
but for successfully coping with increasingly complex human social systems
(Humphrey, 1984; Byrne et al., 1988; Dunbar and Shultz, 2007).
Yet, it is becoming increasingly clear that socialaffective processing
in the human brain is probably realized by domain-general
brain regions and networks not specific to maintaining social interactions
(Bzdok et al., 2015 "Neurobiology of Morality"; Behrens et al., 2009 Computation;
Barret et al., 2013).
These considerations entail a second challenge in brain science:
It is probably impossible to know what purpose neural
processing in a given part of the brain has originally evolved to serve.
We can only observe external manifestations and correlative relationships of
this latent biological purpose.


Importantly,
no two human brains are alike.
Quite the opposite,
they differ with regard to
the morphology of gyri and sulci,
the topology of cytoarchitectonically and chemoarchitectonically
distinguishable brain areas,
the axonal connections linking these brain areas,
as well as the history of their sensory inputs.
%
The extent of a brain area and its inter-individual variability
can be quantitatively examined with its relation to cognition and behavior,
that is,
performance in psychological tasks in the healthy or diseased brain.
For instance, the volume of the amygdala is linked to
interindividual differences in memory performance as well as
many other (temporally transient) states and
(temporally enduring) traits.
%
As another challenge in brain science,
it is currently unknown how
brain-behavior correlations are mediated.
%
The renowed neuroanatomist Santiago Ram\'{o}n y Cajal wrote (1909):
"The complexity of the nervous system is so great,
its various association systems and cell masses so numerous and
complex, and challenging,
that understanding will forever lie beyond our most committed efforts."
%
More specifically, it remains largely elusive
whether distinct behavioral differences
between individuals are associated with changes of
cell bodies, dendrites, axonal connections, and/or glial cells
(Kanai et al., 2011).
That is,
we do not have clear understanding of how
this set of microstructures interact to
solving neural computation problems,
let alone their inter-individual differences.
%
From a methodological perspective,
the conventionally employed
volumetric modelling techniques are na\"ive to
many types of possible morphological differences.
For instance,
it is currently difficult to statistically grasp
inversely proportional left and right hemisphere volumes
or
a condition that randomly affects either the left or the right brain
per individual
(Ashburner et al., 2011).


Worth to be proposed as an independent challenge of
brain science, the secret of interhemispheric
asymmetry is yet to be unveiled.
The connectivity differences between the left and right brain are
for instance currently underresearched.
They are even hardly known in the monkey (Stephan, 2007)
that usually serves a fallback system for human
connectivity investigations (Mesulam, 2012 "The evolving").
In humans, the majority of homologous brain areas feature
direct anatomical connections.
Nevertheless, as two textbook examples,
why the language and attention processes typically lateralize to
the left and right hemisphere, respectively,
is currently understood only in fragments
(Corbetta 2000; Stephan et al., 2003; Price et al., 2010).


It is further unlikely that we
will reach exhaustive understanding of
the human brain by mere
\textit{observational}, as opposed to \textit{interventional},
classes of research methods.
This idea is reflected in
Edward O. Wilson's word s
"disturb Nature and see if she reveals a secret"
as well as in
G. M. Shepherd's words
"Nothing in neuroscience makes sense except in the light of behavior."
%
Purposely induced focal lesions of brain tissue in rats have early
been systemically related to resulting differences in
behavioral performance indices
(Franz and Lashley, 1917).
In hamsters, cats, and monkeys,
decortication entails only small sensory or motor effects,
while such tissue impairments of the neocortex in humans
result in much more pronounced and less reversible 
functional deficits than similar lesions in humans
(Lashley, 1952; MacLean, 1982),
which points to increasing corticalization of brain function.
In humans, brain lesion studies and direct brain stimulation during
neurosurgical interventions have been
the most common approaches
to localize brain functions
until about 20 years ago.
%
However, inferring neurobiological insight from lesion findings constitutes
yet another challenge to brain science.
It constitute an overly simplistic conclusion that
changes in behavior after destroying brain tissue in a
circumscribed brain area directly reveals functional roles
of that brain area (Young, 2000).
%
First,
the destroyed brain area might  primarily subserve inhibitory effects,
such that abolition can increase neural processing subserved in remote areas
mediated by network connections.
Second,
a large fraction of human lesion cases are stroke patients.
The spectrum of lesion patterns found in these populations
is however seriously limited by the existing spectrum of
brain vessel anatomy.
Third,
there is probably not a single psychiatric disorder that would be
characterized by very \textit{focal} (as opposed to distributed)
differences in brain structure
(cf. Goodkind, 2015 JAMA).
%
More generally, it is still a matter of debate whether
structure (i.e., locally specific micro- and chemoarchitecture),
connectivity (i.e., short- and long-rang axonal targets),
and function (i.e., lesion-induced behavioral changes)
reflect three  viewpoints on the same heterogeneity of
a particular brain area
(Passingham et al., 2002; Kelly et al., 2012 Neuroimage).


Each area in the brain exhibits
activation patterns of neuronal populations with
oscillatory regularities.
These oscillatory circles and their associated behaviors are
highly preserved in mammalian evolution
(Buzsaki, 2013 "Scaling brain size").
%
Perhaps since Hubel and Wiesel's (1965) description of increasingly
complex processing of neurons in the primary visual cortex
neuroscientists tend to think information processing as
serial sequences of incoming bottom-up sensory and
modulating top-down information streams.
Axonal feed-forward and feed-back connections
are indeed a very good predictor of \textit{what} the next processing step is.
Yet, brain oscillations are capable of predicting \textit{when}
this next processing step will occur.
%
Oscillation measured by EEG and MEG techniques might be the most
attractive access to another challenge in brain science:
\textit{the binding problem} (Singer, 1999; Engel, 2001; Varela, 2001).
We are far from understanding how
environmental perturbation by multi-sensory stimulation
is coherently integrated and linked with prior experience into
a holistic higher-order percept via
spatially distributed and temporally coherent electrophysiological activity.
%
In animals, oscillatory but not spiking activity of neuronal populations appears
to be closely associated with sensory input processing.
The interpretation of oscillation findings is however demanding.
This is because they
simultaneously reflect
a maintenance equilibrium, sensitivity to external stimuli, and
formation of processing outputs.
For instance,
perception is an intrinsically probabilistic process with nonidentical
results depending on the state of ongoing oscillatory circles.
Additionally,
different "rhythms" (i.e., frequency bands) flank each other
in a same brain area in an interacting fashion.
The same rythm can reflect different categories of computational processes
in different brain areas and networks. Some brain structures are
characterized by specific rythms that may not be found in the rest
of the brain.
Different frequency bands can subserve a same cognitive process, while
different cognitive processes can be realized by the same frequency bands.
%
Finally,
high frequencies govern large-scale networks in the brain that, in turn,
influence small local neuronal space with slow oscillatory patterns.


Also from a philosophical perspective
the neuroscientist faces problems when articulating observations of
phenomena in the brain.
%
For instance, brain areas or experimental effects are frequently
described according to "emotional" versus "cognitive"
interpretational categories.
However, this class of judgments implicitly preassumes
the neurobiological validity of traditional psychological categories.
That is, it assumes that those two concepts have
a discrete representation in measurable neurobiology.
Yet, as another major challenge to brain science,
it remains elusive how and to what extent psychological terms,
such as "emotion" and "cognition" (Pessoa, 2008; Van Overwalle, 2011),
map onto regional brain responses
(Laird et al., 2009; Mesulam, 1998; Poldrack, 2006). 
Potentially unjustified a-priori hypotheses are imposed
on the organization of human brain systems.
%
It should hence be carefully called into question what terms
are an adequate word choice to refer to discrete
neurobiological processes.
More globally,
confusion introduced by human language itself is at the origin
of many scientific problems
(Wittgenstein, 1953/2001 "Philosophical Investigations").
The grammatical and lexical constraints of human language might be
too tight to allow for unequivocal description of the diverse
circumstances humans encounter in science and ever-day life.
According to Wittgenstein the meaning of language is primarily
defined by its practical use in concrete situations,
rather than decontextualized abstractions necessarily pre-shared
by the interlocutors.
Words might not have an objective meaning
equally accessible to and understood by everybody.
This is all the more the case for language descriptions
of phenomena that do not occur in every-day reality.
Discussing subtleties of abstract neurobiological concepts,
which can hardly be practically experienced,
are frequent subject to ambiguity, thus leading to
unnoticed misunderstanding and unresolvable paradoxes
(cf. Bostrom, 2002; Watzlawick et al., 1967).
Biological processes in the brain are an instance of
such not directly experienceable phenomena underdetermined
by human language that entail interpretative conundrum.
%
More concretely,
there is still no community-wide consensus on
a comprehensive description system of
human mental operations (Poldrack, 2006 and 2011).
This has caused considerable heterogeneity in
how neuroimaging experiments have been motivated and conducted.
Moreover, it resulted in frequently inconsistent findings
that are difficult to reconcile conceptually. 
%
In sum,
cognitive neuroscience has so far heavily relied on concepts
borrowed more traditional, non-neurobiological scientific disciplines.
These considerations are especially relevant to investigations
whose conclusion heavily rely on CS.
Statistical hypothesis testing makes the
strong implicit assumption that
the semantics concepts used to forumulate 
the null and altenrative hypotheses are "true".


The last challenges to the neuroscientist
mentioned here are of epistemological origin.
Biology as a whole has modest legacy in abstract theory.
This probably includes the history of the biology of the brain.
In particular,
the spectrum of permissible conclusions
that can be drawn from neuroscientific investigations is strongly conditioned
by the following three questions
(Carruthers, 2009; Dehaene, MBE 2007):
\begin{enumerate}
  \item Does the human brain offer sufficient computational resources
  to grasp, formalize, and predict itself?
  \item Is the human mind capable to reflect upon itself
by directly contemplating itself (i.e., introspection) or by
indirectly contemplating an internalized self-model acquired through
interaction with others?
  \item To what extent is the self-reflexive description of the phenomenology of
  the human mind by the human mind itself immanently
  limited and paradoxical?
\end{enumerate}


Taken together,
there are many intricacies about neurobiology and 
the mosaic knowledge that we currently have about it.
%
Despite $\approx$200 years of neuroscience,
we are probably not even close to something like a unified theory of
brain function
that neuroscientists from different fields would accept
(cf. Friston, 2010 "Free energy principle"; Bar, 2009 "Predictions").
This caveat considerably complicates
the formulation of precise, neurobiologically
valid hypotheses that can be experimentally tested in targeted studies.
%
Therefore, it might be helpful to use heuristics-establishing
statistical approaches for pattern discovery instead of
classical statistics alone.
%
Discovering the mystery of the brain \textit{exclusively} by
successive falsification of
entirely human-conceived,
intimately language-dependent,
and dichotomically framed hypotheses
might be a viewed as hybris by some
(cf. Cajal, 1909; Cohen, 1994).
%
Therefore,
the present dissertation is built on the assumption that
we might not reach an \textit{exhaustive analytical understanding}
of the brain any time soon
and that a more pragmatic access
may rely in the \textit{heuristic
approximation of brain mechanisms} by statistical learning models.
%
Such an attempt to learn patterns from data
would follow the same
direction as recent research developments in
language translation and drug discovery (cf. 1.1).



\subsection*{1.4 The curse of dimensionality}

Not only conceptual and neurobiological challenges,
but also the increasing quantities of analyzed data
put neuroscientific research to the test
(Gorgolewski \& Poldrack, NNR).
%
Today's neuroimaging methods offer very high resolution in
space (especially fMRI and PET)
and
time (especially EEG and MEG)
(Amunts et al., 2014 Science; Buzsaki \& Draguhn, 2009 Science).
%
The mere number of features poses serious
statistical challenges to the investigator.
It is the neuroscientific version of what Richard Bellman
called the \textit{curse of dimensionality} (1961).
%
At the root of the problem,
all data samples look virtually identical
in high-dimensional data scenarios.
%
Accustomed to regularities in 3D neighborhoods,
human intuition is often led astray in
how data behave in
input spaces with an extreme number of variables.



The more dimensions an input space spans,
the further the data points are away from each other
(Hastie et al., 2011).
Conter-intuitively,
measuring the distance between a randomly selected data point
and its closest uniformly distributed neighbors,
reveals a shell-like occurrence probability of
these neighbors, rather than a centered probability mass.
%
Put differently,
when approximating a hypersphere by a surrounding hypercube,
the probability mass of the hypercube
would almost entirely lie outside the hypersphere
(Domingos, 2012).
%
Put in yet another way,
a space divided into isotropic units grows exponentially in the
unit number with linearly increasing dimensionality.
As the main practical conclusion,
the amount of data necessary to populate these units
also grows exponentially
with linearly increasing inputs variables
(Bishop, PRML).



Additionally,
the target function is almost always unknown
in statistical learning investigations.
Hence, we frequently have no knowledge of whether or not
special structure may exist in the input data that can be exploited.
%
Knowledge of special structure of the phenomenon under study
can reduce both \textit{bias}
(i.e., difference between the target function and
the average of the function space derivable from a model)
and
\textit{variance}
(i.e., difference between the best approximating function 
from the function space and
the average of the function space).
This is a rare opportunity in SL because increasing,
for instance, the model complexity
typically increases the variance and lowers the bias, and vice versa.
%
In particular,
the problem of overfitting in SL has an immediate relationship
with the multiple-comparisons problem in CS
(Domingos, 2012).
%
The \textit{bias-variance decomposition} captures the fundamental
tradeoff in statistical modeling between
approximating the behavior of
the studied phenomenon and
generalizing to newly generated data describing that behavior. 



A peacefully coexisting conceptual framework exists in SL
that is independent
of the unknown target function.
The \textit{Vapnik-Chervonenkis (VC) dimensions}
formalize the circumstances
under which learning processes can be successful (Vapnik, 1989, 1996).
This comprises any instance of learning
from a number of observations
to derive heuristic rules that capture properties of phenomena in nature,
including learning in humans and machines.
Formally, the VC dimensions measure the
complexity capacity of a class of approximating functions
(i.e., the function space). 
%
Practically, good models have finite VC dimensions
and are therefore capable to generalize to new data.
Bad models have infinite VC dimensions that
are unable to make generalization conclusions on unseen data,
regardless of data quantity.



More concretely,
SL approaches that incorporate locally varying functions
in small \textit{isotropic} neighborhoods
will fail to generalize in high-dimensional data scenarios.
SL approaches that overcome the curse of dimensionality typically
incorporate an explicit or implicit metric for
\textit{anisotropic} neighborhoods
(Hastie et al., 2011).
%
It is the \textit{hyperparameters} that govern the
smoothing behavior of the imposed local neighborhoods.
%
In so doing,
the \textit{hypothesis set} (i.e., each function in the function space
represent a hypothetical solution to
the estimation problem) is hopefully reduced to
a reasonable pre-selection (i.e., \textit{regularization}).
%
Guiding the statistical estimation process by
complexity restrictions can alleviate the curse of dimensionality.
First,
we can deliberately exclude members of the hypothesis set.
Viewed from the bias-variance trade-off, this calibrates
the sweet spot between underfitting and overfitting.
Viewed from Vapnik's statistical learning theory,
the VC dimensions can be reduced and thus the generalization performance
increased.
%
Second, there is an infinity of possiblities to restrict the hypothesis set.
Yet, these choices are typically guided by external knowledge beyond
the data at hand.
%
Third,
different complexity restrictions typically lead to different
best approximating functions.


In sum,
the choice of any statistical method contraints
the spectrum of possible results and of permissible interpretations.
Any scientific discovery in the brain is only valid in the
context of the complexity restrictions that have been imposed
on the neurobiological phenomenon of interest.
%
No single statistical strategy, be it SL, CS, or other,
can consistently
do better in all neuroscientific investigations
(Wolpert, 1996).
%
The present dissertation
is hence dedicated to the juggling with
complexity restrictions to
neurobiological reality as observed by fMRI scanning.



\subsection*{1.5 Imaging neuroscience}



- Argumente von DFG Antrag

Until 20 years ago, lesion studies and brain stimulation during neurosurgery were the main tools to investigate human brain function. The neuroimaging era of positron emission tomography (PET) and functional magnetic resonance imaging (fMRI) then enabled non-invasive, in-vivo brain research in health and disease. Experimental neuroimaging has provided insight into the cerebral localization of sensory processing, motor actions, and affective functions. Yet, the working principles of the human brain still remain largely elusive.


we have to elucidate the regionally specific localization of mental processes. Until about 20 years ago, the most common approaches to localize human brain functions consisted in lesion studies and direct brain stimulation during neurosurgical interventions. The beginning of the functional neuroimaging era by positron emission tomography (PET) and functional magnetic resonance imaging (fMRI) then enabled the non-invasive, in vivo investigation of functional specialization in the human brain. Based on local changes in cerebral blood flow, glucose- or oxygen-metabolism, these techniques allow the identification of regional increases in neural activation during the performance of specific tasks. Usually, the neural correlates of a given process of interest are isolated by subtraction of the activation measured a closely related (control) task that is supposed not to evoke the psychological process of interest. Evidently, the spectrum of neuroimaging-compatible tasks is almost only limited by the scanner surroundings and the interdiction of head movements. In fact, the constantly increasing publication rate currently counts more than 1,000 new neuroimaging articles per year [1]. Functional neuroimaging has hence provided a wealth of information on the cerebral localization of sensory processing, motor actions as well as cognitive and affective functions.

In spite of its success in revealing the location of regionally specific effects in health and disease, however, functional neuroimaging using PET or fMRI suffer from several limitations that restrict the amount of knowledge that may be gained from each experiment:
Usually, between 15 and 30 subjects participate in a neuroimaging study. Although this denotes a remarkable increase since the beginnings of neuroimaging science, it is still a rather small sample size comparing to other scientific disciplines, such as social sciences, population genetics or clinical trials. In contrast to experimental psychology, the considerable logistical and financial efforts of each neuroimaging study discourage replication studies as well as the combination of several studies, leading at times to the publication of isolated findings.

the recorded signals are thus potentially confounded by a vast range of biological, technical and methodological factors, which reduce reliability of the obtained results. In fact, the underlying relationship between neuronal activity and measured signals is currently not exhaustively understood. On the other hand, PET and fMRI yield relative, rather than absolute, signals related to neuronal activity. The neural correlates underlying a given psychological task thus need to be derived by subtraction between two experimental conditions rendering the resulting activation pattern context-dependent. It has, however, repeatedly been questioned that conclusions on general physiological or pathological mechanisms can be drawn if reliant on the neuroimaging-inherent subtraction logic [12]. In other words, even though frequently discussed in terms of general mechanisms, the results of a particular neuroimaging study actually just pertain to the difference of the two specifically chosen conditions.

These manifold limitations of any neuroimaging study thus severely impede the amount of knowledge that may be gained from each individual neuroimaging study. This consideration has consequently prompted a need for synthesizing and integrating existing research. By judging concordance and disconcordance across a high number of studies, the limitation of each individual PET or fMRI experiment may be overcome and a synoptic view on earlier isolated findings can be acquired. 

That is, most neuroimaging articles report significant activation foci according to either the Talairach-Tournoux [13] or Montreal Neurological Institute [MNI; 5] 3D reference space, which in turn are easily being converted into each other [8]

although the mechanisms underlying the information produced are not fully understood (Logothetis, Pauls, Augath, Trinath, & Oeltermann, 2001).


Taken together, the biggest fraction of the various brain signals does not correlate with a particular behavior, stimulus, or experimental task. The constant blood flow, oxygen intake, and energy consumption are highly linked. These partially uncouple in a task setting, but the relative change is very small.

these techniques allow the identification of regional increases in neural activation during the performance of specific tasks. Usually, the neural correlates of a given task (reflecting a mental process of interest) are isolated by subtraction of the activation measured during a closely related task (i.e., control task) that is supposed not to evoke the psychological process of interest. Evidently, the spectrum of neuroimaging-compatible tasks is practically only limited by the scanner surroundings and the interdiction of head movements.


Some properties:

highly vascularized cortical areas (e.g., primary sensory areas and the speech-related Broca and Wernicke's areas) have a higher baseline probability to be observed in neuroimaging studies (Harrison et al., 2002).

Blood demand and blood supply generally work in concert in the human brain. In the example of task-induced metabolic consumption, onset of vibratory stimulation of participants' hand entails regional accumulation of metabolic equivalents that cause regional blood flow increase (cf. neurovascular coupling below) in the contralateral somatosensory cortex (Fox et al., 1986). On a microanatomical scale, local increases in neural activation entail privileged local blood supply at the expense of immediately neighboring neuronal populations (i.e., vascular-steal phenomenon). On a macroanatomical scale, however, regional increases in brain activity due to recruitment of a particular cognitive process induce regional blood flow changes of only a few percent. The regional changes in blood flow accompanied by cognitive phenomena are thus not measurable as changes in total blood flow to the brain, although the brain's hemodynamic reserve permits doubling the blood influx.

The neuroscientist is typically interested in neural activity, which was suggested to induce regional blood flow changes already in the 19th century (Roy et al., 1890). Physiologically, neurovascular coupling describes the brain vasculature's time-lagged dilation response to locally increasing metabolic products. This is directly observable using water or butanol positron emission tomography (PET) and indirectly using functional magnetic resonance imaging (fMRI) as the BOLD (blood-oxygen level-dependent) signal (Fig. 1). The BOLD signal is caused by local changes in blood oxygenation, i.e., the oxyhemoglobin versus deoxyhemoglobin ratio (red blood cells carrying versus not carrying oxygen) (Ogawa et al., 1990; Logothetis et al., 2004). It exhibits an initial dip after the onset of neural activity increase that is attributed to the fast local increase in deoxyhemoglobin. The ensuing hyperperfusion and the thus generated (relative) hyperoxygenation (cf. next paragraph) then dictate the BOLD signal shape (hemodynamic response function). It is variable across the brain regions of an identical organism, across organisms, and probably across different tasks. Neural activation is finally followed by re-inhibition of blood flow observable as an undershoot at the end of the BOLD signal (Logothetis et al., 2001).
Comparing neural activity and corresponding BOLD signals, the BOLD signal is at least one order of magnitude noisier, scales roughly linearly with neural activity, and is better predicted by local field potentials than multi-unit spiking activity. The BOLD signal is possibly more associated with input to and processing in a local neuronal population rather than its output. There is thus no clear-cut quantitative relation between the spike rate of neuronal populations and the ensuing BOLD response. Rather, the BOLD signal reflects a mixture of transient spikes and continuous membrane potentials (Logothetis et al., 2004).

This implies intra- and inter-neuronal house keeping functions as particularly important in the human brain's resting functionality. In contrast to this baseline consumption, environmental events evoke increased energy costs of only less than 5%. Importantly, however, most neuroimaging investigations focus on these (minor) metabolic changes due to external stimulation, rather than the majority of metabolic demands due to biochemical maintenance of neuronal architecture.









- Their popularity of SL in neuroimaging increased dramatically in the attempt of "mind-reading" or "decoding" cognitive processes from neural activity patterns (Haynes and Rees, 2005; Kamitani and Tong, 2005)


neuroimaging: nature of the atomic neurobiological substrates
that increase or decreases is subject to debate.



It has been argued that classical functional neuroimaging applications
are insufficient to infer structure-function relationships
without formal modelling (Stephan, 2004).
(e.g. Flechsig 1905; Meynert 1890)
-> towards formalized descriptions of SFRs


This dissertation is concerned with one of the findings that
we might never have discovered with the emergence of
brain imaging techniques:

DMN:

Without focus on an explicit task the brain is not at rest.

On a macro-scale, the metabolic baseline turnover is not equally distributed across the brain. The metabolic demands are variable across brain regions and (to a smaller extent) across time. The highest metabolic consumption locates, first, to the posterior cingulate cortex extending into the adjacent retrosplenial cortex and precuneus and, second, to the medial prefrontal cortex extending into the anterior cingulate cortex (Raichle et al., 2001; Reivich et al., 1979).

Resting-state signal: Fluctuations in physiological signals recorded in the absence of externally structured experimental settings that are reflected in voxels' time courses. These signals are thought to have neuronal, metabolic, cardiac, and respiratory origins.

Resting-state network (RSN):  Coherent spatiotemporal coupling in resting-state BOLD fluctuations yields a set of robust neural networks.

Taken together, the biggest fraction of the various brain signals does not correlate with a particular behavior, stimulus, or experimental task. The constant blood flow, oxygen intake, and energy consumption are highly linked. These partially uncouple in a task setting, but the relative change is very small.

In the absence of task (i.e., during mind wandering), inter-neuronal communication continues.

Importantly, the (small) amplitude of the RS signal is modulated by transient psychological states (e.g., arousal, attention, and alertness) as well as cardiac and respiratory influences. Indeed, the decomposability of the RS signal into independent components suggests a set of distinct influences rather than one coherent signal pattern (Fukunaga et al., 2006). More specifically, evidence exists in favor of a neuron-, metabolism-, vasculature-, and oxygen-driven genesis of the RS signal. 

To date, the stratification of the physiological factors underlying the RS signal is poorly understood. It is, however, commonly agreed that the variability in the RS signal arises from both physiological factors and the organism's mental operations.

It is becoming increasingly clear that brain areas pertaining to social cognition are topographically highly similar to brain areas that increase activity in the idling, unconstrained mind, the so-called “default mode”, and decrease activity during stimulus-driven, goal-directed tasks (Shulman et al., 1997; Spreng et al., 2009). More specifically, brain areas underlying unconstrained cognition were consistently associated with a number of complex, introspective mental tasks, including contemplating mind states, self-focused reflection, mental navigation of the body in space, autobiographical memory recall, and, more generally, envisioning situations detached from reality. 

In task-based neuroimaging studies, the neural correlates underlying a certain psychological function are derived by subtraction between two conditions belonging to a so-called experimental paradigm [5, 6]. The neural correlates of a mental process of interest are thus isolated by subtraction of the activation measured during a closely related task (i.e., control task) that is supposed not to evoke the mental process of interest. Despite the many merits of experimental neuroimaging, it has repeatedly been questioned, from a scientific perspective, whether conclusions on healthy neurophysiology or pathological mechanisms can be drawn if relying on the subtraction logic inherent to experimental neuroimaging [7]. As the results of a particular experimental neuroimaging study reflect the difference of two specifically chosen conditions, similarities and dissimilarities observed across psychiatric disorders might be entirely epiphenomenal. From a more clinical perspective, task-based neuroimaging is also highly unlikely to become diagnostic routine in psychiatry given the necessary human (e.g., importance of administering an experimental paradigm) and hardware (e.g., stimulation devices and presentation programs) infrastructure as well as the numerous manual choices (i.e., subjective interventions) during data analysis.

With respect to structural neuroimaging, two methods are currently dominating (clinical) systems neuroscience. In voxel-based morphometry (VBM) studies, a T1-weighted high-resolution MRI scan is acquired for each participant [8, 9]. These are then automatically segmented into the cerebrospinal fluid, grey (GM) and white (WM) matter. After warping each of these structural images into a common template image in stereotaxic standard space, GM and/or WM are submitted to statistical analysis. This allows for quantitative comparison of the regional amount of GM and WM volume [10], while taking into account individual brain sizes. Diffusion tensor imaging (DTI) studies, on the other hand, do not assess voxel-wise volumetric information but the principal direction, and more specifically also the directedness, of white-matter fiber bundles at each location [11, 12]. This is achieved by measuring to what extent the random walk of local water molecules under diffusion is rather spherical (i.e., moving freely in space) or rather ellipsoid (i.e., moving constrained by axon-sheath-like boundaries). The voxel-wise information on the directedness of diffusion as a proxy for fiber architecture can then be compared between subjects (commonly by fractional anisotropy [FA] as a summary measure) or used to delineate the three-dimensional trajectories of fiber tracts in the human brain in-vivo.
In comparison to task-based neuroimaging studies, both VBM and DTI studies do not require the participants' adherence to an experimental paradigm. This is of tremendous advantage especially in psychiatric populations given that disturbance of thought defines these medical conditions. It further entails that task-free VBM and DTI studies are not only less demanding and more replicable but also yield easier, faster, and more standardized results. Structural neuroimaging studies, compatible with most standard clinical scanners, thus appear much better suited for clinical translation than task-based investigations. However, VBM and DTI do not yield any information on dynamic brain features, such as neural activity strength, inter-regional coupling, or recruitment of functional brain modules. This represents a major drawback as functional disintegration rather than structural abnormality is currently believed to be at the core of many psychiatric disorders. Additionally, structural neuroimaging markers appear to be more confounded by non-clinical factors (especially age) than functional data.




Correlation analysis can detect temporal coincidence in spontaneous, slow fluctuations (~0.01 - 0.1 Hz) in fMRI signals. This is taken as a measure of RS connectivity between topographically distant parts of the brain. Faster frequencies, not apparent in (intrinsically low-pass filtered) BOLD signals, in turn are predominantly associated with cardiac and respiratory sources. RS connectivity is task-unconstrained by probing participants that lie in the scanner without following a defined experimental task. Instead, participants are instructed to think of nothing in particular, let their minds go, and leave their eyes open/closed or look at a fixation cross. During mind wandering humans mentally shift between various heterogeneous types of thoughts, memories, and predictions. This is why RS connectivity is believed to reflect the repertoire of cognitive operations that the human brain can perform (Smith et al., 2009).


Observed RS connectivity patterns led to discovery of the coherent spatiotemporal fluctuations termed resting-state networks (RSN).

The probably best-known RSN is the default-mode network (DMN) (Gusnard et al., 2001) that was discovered with and mainly characterized by RS connectivity. The DMN was initially proposed to be exclusive in decreasing neural activity consistently during experimental paradigms requiring stimulus-guided behavior. This RSN was therefore proposed to reflect the neural correlates underlying unfocused every-day mind wandering (Schacter et al., 2007). This, in turn, would suggest a close correspondence between the physiological and psychological baseline of the human brain (Schilbach et al., 2008). It was later even argued that this network is systematically anti-correlated with brain regions more active during task performance (Fox, et al., 2005). This was recently challenged by repeated reports of brain regions exhibiting both task-constrained and task-unconstrained increases in neural activity (Buckner, et al., 2008). More specifically, the DMN is now known to consistently increase neural activity during a small set of complex cognitive tasks, including social cognition, autobiographical recall, and spatial navigation (Bzdok, et al., 2013; Spreng, et al., 2009). Task-, RS- and meta-analysis-informed research on the DMN thus corroborated that this particular network is consistently decreases activity during externally focused mental tasks and typically increases activity during a small set of internally focused mental tasks.



It is a limitation of these studies that they attempt to derive the normal function of an area from the effects of damage to that area.

On a macroscale, the metabolic baseline turnover is not equally distributed across the brain. The metabolic demands are variable across brain regions and (to a smaller extent) across time. The highest metabolic consumption locates, first, to the posterior cingulate cortex extending into the adjacent retrosplenial cortex and precuneus and, second, to the medial prefrontal cortex extending into the anterior cingulate cortex (Raichle et al., 2001; Reivich et al., 1979).
-> but we have no idea why


example of serendipity / A property of the brain that we might not have
discovered without the advent of brain imaging methods.

The default-mode network - we are not even close to knowing
what this baseline/house-keeping activity network is doing

A resting-state connectivity derived network that was initially proposed to be consistently decreasing neural activity during experimental paradigms requiring stimulus-guided behavior. It is now believed to be mostly, but not exclusively, active in the absence of external stimulation. 
15 years ago neuroscientists discovered a unique brain network. The so-called default-mode network (DMN) appeared to be exclusive in decreasing neural activity consistently during experimental tasks.
It was later even argued that this network is systematically anti-correlated with brain regions active during task performance. This was recently challenged by repeated reports of brain regions exhibiting both task-constrained and task-unconstrained increases in neural activity. On the one hand, the DMN is now believed to be mostly, but not exclusively, active during absent external stimulation. On the other hand, the DMN is today known to also increase neural activity consistently during a small set of complex cognitive tasks. This includes the contemplation of others’ and one’s own mind states, as well as scene construction processes when envisioning past, fictitious, and future events. It was speculated that the human brain might have evolved to, by default, predict environmental events using mental imagery. Constructing detached probabilistic scenes could thus influence perception and behavior by estimating saliency and action outcomes. This would invigorate a possible relationship between the physiological baseline of the human brain and an introspective psychological baseline.
The DMN defies our neuroscientific intuitions and challenges established methodology. This brain network likely represents a physiological instantiation of a human beings' default mental repertoire, the nature of which remains largely obscure.

This functional segregation is underlined by the here observed anti-correlation between aRTPJ and pRTPJ. Please note that the repeatedly shown anti-correlation between the (aRTPJ-related) saliency and (pRTPJ-related) default-mode network (Fox et al., 2005a; Kelly et al., 2008) was here derived from, and is thus directly related to, different functional modules within RTPJ. Indeed, goal-directed task performance improves with increased activity in saliency-related areas and decreased activity in default-mode areas (Weissman et al., 2006). Conversely, increased activity in areas of the here delineated pRTPJ network were linked to increased occurrence of task-independent thoughts (i.e., mind-wandering) during task execution (Mason et al., 2007). Two fMRI studies employing Granger causality analysis further corroborated the anti-correlation by indicating negative influence of the default-mode on the saliency network (Pisapia et al., 2012) and vice versa (Sridharan et al., 2008). Taken together, present and previous findings converge on the tentative notion that the human RTPJ contains two functionally complementary modules that belong to antagonistic networks potentially underlying externally versus internally oriented processing.

The probably best-known RSN is the default-mode network (DMN) (Gusnard et al., 2001) that was discovered with and mainly characterized by RS connectivity. The DMN was initially proposed to be exclusive in decreasing neural activity consistently during experimental paradigms requiring stimulus-guided behavior. This RSN was therefore proposed to reflect the neural correlates underlying unfocused every-day mind wandering (Schacter et al., 2007). This, in turn, would suggest a close correspondence between the physiological and psychological baseline of the human brain (Schilbach et al., 2008). It was later even argued that this network is systematically anti-correlated with brain regions more active during task performance (Fox, et al., 2005). This was recently challenged by repeated reports of brain regions exhibiting both task-constrained and task-unconstrained increases in neural activity (Buckner, et al., 2008). More specifically, the DMN is now known to consistently increase neural activity during a small set of complex cognitive tasks, including social cognition, autobiographical recall, and spatial navigation (Bzdok, et al., 2013; Spreng, et al., 2009). Task-, RS- and meta-analysis-informed research on the DMN thus corroborated that this particular network is consistently decreases activity during externally focused mental tasks and typically increases activity during a small set of internally focused mental tasks.

Mental imagery provides one of at least three settings where the activation of sensory areas can transcend the constraints imposed by external reality. The two other examples include working memory, where the representation of a sensory experience is prolonged beyond the duration of the stimulus, and anticipatory paired-associate responses, where sensory representations become activated in expectation of an event which is not yet present in reality. (MES)




\subsection*{1.6 Statistical learning approaches in brain imaging}

HCP / TERA

- Classification approaches were early used in structural neuroimaging (Herndon et al., 1996) and contributed to improved image preprocessing performance (Ashburner and Friston, 2005).



Much of the success of cognitive neuroscience over the past decades has been a result of implementing a mass-univariate analysis of neuroimaging data obtained from human subjects as part of a general linear model (GLM). GLM treats each volumetric pixel, i.e. voxel, as independently to perform serial univariate statistics (Friston et al., 1995). 
Univariate approaches are recognized to be an excellent approximation for topographically localizing activations, i.e. a differential increase of neural activity, in individual voxels. Multivariate (ML) approaches on the other hand can be used to examine responses that are jointly encoded in multiple voxels. In other words, a univariate statistical model considers a single voxel at a time, while a multivariate model considers many voxels simultaneously and can make inferences on distributed responses without requiring strong independence assumptions. 

Statistical methods with non-parametric properties, often summarized as "multivariate pattern analysis" (MVPA), have recently gained popularity in neuroimaging. In particular, ML approaches are well suited for discovering complex structure between high-dimensional neuroimaging data (mostly functional magnetic resonance imaging) and variables of interest (e.g. external stimuli or experimentally imposed cognitive sets). 


For decades histopathology, lesion studies, and invasive research in non-human primates have been the main neuroscientific workhorses. The advent of neuroimaging methods then leveraged unprecedented insight into brain biology. Task-based neuroimaging studies typically attempt to solicit target mental operations in participants by means of sensory, cognitive, or affective paradigms. This research greatly increased our understanding of the healthy and diseased human brain. Yet, such work rests on a set of strict assumptions, including that a) historically inherited psychological categories are useful description systems for neurobiological phenomena, b) target and non-target mental operations can be statistically distinguished by so-called “cognitive subtraction,” and c) a few dozen participants are sufficient to reliably establish fundamental (patho-)physiological mechanisms in the brain. All these assumptions have repeatedly been questioned. Apart from that, structural neuroimaging approaches, such as voxel-based morphometry, may be used to identify locations in which two groups of subjects, such as patients versus controls, differ from each other in local gray matter (GM) volume. While not being affected by the choice of a particular task, these approaches are also frequently criticized for insufficient sample sizes. In addition, they only allow characterizing presence, absence, or a group-mean effect at a particular location. Finally, methods for individual structural analyses are manifold and have repeatedly been in disagreement.

As a complementary methodological family, machine learning (ML) approaches are characterized by a) making the least assumptions possible, b) being more motivated by mathematical models rather than cognitive theory, and c) automatically mining structured knowledge from massive data resources. Given the widely acknowledged intricacies of and slow progress in schizophrenia research, ML methods lend themselves particularly well. They have the potential to decipher and subsequently render explicit biological indices that might assist diagnosis, treatment, and clinical outcome prediction.


Typical big-data problems concern classification or regression of an output variable y with respect to a large number of input parameters x, also called predictor variables or covariates, on the basis of large training sets. 


Note that this curse of dimensionality does not automatically apply to all big-data algorithms. To the contrary, it occasionally turns out helpful to artificially increase the dimensionality of the parameter space in methods like decision trees or support vector machines 

- Their popularity of SL in neuroimaging increased dramatically in the attempt of "mind-reading" or "decoding" cognitive processes from neural activity patterns (Haynes and Rees, 2005; Kamitani and Tong, 2005)


Third, regarding data analysis techniques, it has been proposed to leverage ML as a particularly powerful tool to investigate the neural correlates of two interacting brains (Konvalinka\& Roepstorff 2012). ML promises to extend the representational agenda of fMRI investigations (i.e., activation-based analysis) and to an informational agenda (i.e., pattern-information-based analysis) (Kriegeskorte et al., 2006; Mur et al., 2009). In particular, multivariate ML approaches are able to address the four questions whether, where, when and how (Brodersen, 2009). ML can provide new evidence whether a given type of information is encoded by neural activity, where a given type of information is neurally encoded, when it is processed/generated/bound and how information is neurally processed in the human brain. Moreover, single-individual fMRI analyses already pose a high-dimensional statistical problem. Performing two-individual fMRI analyses thus imposes an even more challenging "curse of dimensionality" (Hastie et al., 2011). Automated statistical learning of robust structure (i.e., informational patterns) in large quantities of (neuroimaging) data is a core feature of ML approaches. In this way, the flexibility and capability of ML algorithms for pattern mining in extensive data resources make it particularly attractive for the statistical mastery of online two-brain interaction.

. This prompted the current drift towards more data-driven (i.e., fewer assumptions), higher-dimensional (i.e., more features per observation) neuroimaging analyses on larger samples of neuroimaging data (i.e., more observations).


Pietsch:
The notion of a qualitative change from hierarchical to horizontal modeling is further corroborated by a concurrent paradigm shift in statistics, which has been described as a transition from parametric to non-parametric modeling (e.g. Russell \& Norvig 2010, Ch. 18.8), from data to algorithmic models (Breiman 2001), or from model-based to model-free approaches.


As argued in the following chapters,
we need to discover
the low-dimensional manifolds that are
embedded within the high-dimensional neuroimaging data.

Helmstaedter 2011 / deep learning


\section*{2 Unsupervised modelling of brain regions}

\subsection*{2.1 Motivation}
\subsection*{2.2 Methodological approach}
\subsection*{2.3 Experimental results}
\subsection*{2.4 Discussion}


\bigskip
\section*{3 Supervised modelling of brain networks}

\subsection*{3.1 Motivation}
\subsection*{3.2 Methodological approach}
\subsection*{3.3 Experimental results}
\subsection*{3.4 Discussion}

\bigskip
\section*{4 Semi-supervised modelling for structure discovery
and structure inference}

\subsection*{4.1 Motivation}
\subsection*{4.2 Methodological approach}
\subsection*{4.3 Experimental results}
\subsection*{4.4 Discussion}


\bigskip
\section*{5 Conclusion}



\bigskip
\section*{6 References}

\bibliographystyle{plainnat}
\bibliography{danilos_endnote_list}

\end{document}
